Covid-19 infection and attributable mortality in residents of UK care homes: Cohort study using electronic records (March-June 2020)

### 2020-07-10 manuscript/manuscript.Rmd
edits

---
title: "Covid-19 infection and attributable mortality in residents of UK care homes: Cohort study using electronic records (March-June 2020)"
# author: "Dutey-Magni, PF, Williams, H, Hayward, A, Shallcross, L"
bibliography: references.bib
csl: elsevier-vancouver.csl
output:
  word_document:
    toc: yes
    reference_docx: template.docx
  html_document:
    df_print: paged
    toc: yes
    code_folding: hide
link-citations: yes
always_allow_html: yes
---

@@

At the peak of the pandemic the number of deaths in care home residents was three-fold higher than the equivalent period in in 2019.[@onsDR2020]  Data from the Office for National Statistics suggest that two-thirds of the 29,393 excess deaths recorded in care home residents between 28th December 2019 and 12th June 2020 (calculated relative to equivalent period in the previous year) involved Covid-19.[@onsDR2020] However, estimating the proportion of care home deaths that can be directly attributed to Covid-19 based on death certifications is challenging because many care home residents with suspected infection were not tested, particularly at the start of the pandemic. Understanding the ratio of direct to indirect deaths in care home residents will be critical to inform public health measures to mitigate the impact of the anticipated second wave of the pandemic, since lockdown measures that prevent infection may have a negative impact on residents by disrupting usual care.

@@

Data were analysed in `R`3·5·0[@RCoreTeam2018] using the `epitool`[@Aragon2020] and `survival`[@therneau2015] libraries. All computer syntax is available on an online repository.[@Dutey2020] Ninety-five percent confidence intervals for proportions and rates were computed from the exact Poisson and binomial limits.

@@

```{r fig_KM_mortality,  fig.height = 6, fig.width = 8, message=F}
KM_mortality_by_covid_status <- survfit(
  Surv(time = start_int, time2 = stop_int, event = death_status) ~ covid_status + outbreak, 
  data = mutate(
    residents_overview, 
    death_status = as.integer(status == "Deceased")
  ))
KM_mortality_plot <- ggsurvplot_facet(
  KM_mortality_by_covid_status,
  mutate(residents_overview, 
         death_status = as.integer(status == "Deceased")  ),
  facet.by  = "covid_status",
  legend.labs = c("Non-outbreak home", "Outbreak home"),
  short.panel.labs	 = T, break.x.by	= 7, 
  break.y.by = .1,
  palette = "aaas", conf.int = T,
  scales = "free", ggtheme = theme_bw(), xlim = c(0, 104), ylim = c(.4, 1),
  xlab  = paste0("Days since ", format(date_min, "%e %b %Y")),
  conf.int.alpha = .15, censor.shape = 124, censor.size = 2, size = 0.5)

@@

This work was supported by the Economic and Social Research Council [grant reference ES/V003887/1].

>>>>

---
title: "Covid-19 infection and attributable mortality in residents of UK care homes: Cohort study using electronic records (March-June 2020)"
# author: "Dutey-Magni, PF, Williams, H, Hayward, A, Shallcross, L"
bibliography: references.bib
csl: elsevier-vancouver.csl
output:
  html_document:
    df_print: paged
    toc: yes
    code_folding: hide
  word_document:
    toc: yes
    reference_docx: template.docx
link-citations: yes
always_allow_html: yes
---

Peter Dutey Magni,^{1} Haydn Williams,^{2} Arnoupe Jhass,^{1} Greta Rait, Harry Hemingway,^{134} Andrew Hayward,^{1} Laura Shallcross^{1}

1 Institute of Health Informatics, University College London, London, UK
2 Four Seasons Health Care Group
3 Health Data Research UK, University College London, London, UK
4 University College London Hospitals NIHR Biomedical Research Centre, London, UK
5 Institute of Epidemiology & Health Care, University College London, London, UK

@@

At the peak of the pandemic the number of deaths in care home residents was three-fold higher than the equivalent period in in 2019.[@onsDR2020]  Data from the Office for National Statistics suggest that two-thirds of the 29,393 excess deaths recorded in care home residents between 28th December 2019 and 12th June 2020 (calculated relative to equivalent period in the previous year) involved Covid-19.[@onsDR2020] However, estimating the proportion of care home deaths that can be directly attributed to Covid-19 based on death certifications is challenging because many care home residents with suspected infection were not tested, particularly at the start of the pandemic. Understanding the ratio of direct to indirect deaths in care home residents will be critical to inform public health measures to mitigate the impact of the anticipated second wave of the pandemic, since lockdown measures[@Hale2020] that prevent infection may have a negative impact on residents by disrupting usual care.

@@

Data were analysed in `R`3·5·0 using the `epitool`[@Aragon2020] and `survival`[@therneau2015] libraries. All computer syntax is available on an online repository.[@Dutey2020] Ninety-five percent confidence intervals for proportions and rates were computed from the exact Poisson and binomial limits.

@@

```{r fig_KM_mortality,  fig.height = 6, fig.width = 8, message=F}
KM_mortality_by_covid_status <- survfit(
  Surv(time = start_int, time2 = stop_int, event = death_status) ~ covid_status + outbreak, 
  data = mutate(
    residents_overview, 
    death_status = as.integer(status == "Deceased")
  ))
KM_mortality_plot <- ggsurvplot_facet(
  KM_mortality_by_covid_status,
  mutate(residents_overview, 
         death_status = as.integer(status == "Deceased")  ),
  facet.by  = "covid_status",
  legend.labs = c("Non-outbreak home", "Outbreak home"),
  short.panel.labs	 = T, break.x.by	= 7, 
  break.y.by = .1,
  palette = "aaas", conf.int = T,
  scales = "free", ggtheme = theme_bw(), xlim = c(0, 104), 
  xlab  = paste0("Days since ", format(date_min, "%e %b %Y")),
  conf.int.alpha = .15, censor.shape = 124, censor.size = 2, size = 0.5)

@@

This work was supported by the Economic and Social Research Council [ES/V003887/1].  This work was also supported by a National Institute for Health Research (NIHR) Clinician Scientist award (CS-2016-007 to L.S.); In-Practice fellowship (NIHR300293 to A.J.); a NIHR School of Primary Care Research fellowship (SPCR-118 to A.J.); and a NIHR In-Practice Fellowship (to M.R.). AH and HH are NIHR Senior Investigators . AH and HH are supported by Health Data Research UK (grant No. LOND1), which is funded by the UK Medical Research Council, Engineering and Physical Sciences Research Council, Economic and Social Research Council, Department of Health and Social Care (England), Chief Scientist Office of the Scottish Government Health and Social Care Directorates, Health and Social Care Research and Development Division (Welsh Government), Public Health Agency (Northern Ireland), British Heart Foundation, Wellcome Trust. HH is funded by the NIHR University College London Hospitals Biomedical Research Centre, The BigData@Heart Consortium, funded by the Innovative Medicines Initiative-2 Joint Undertaking under grant agreement No. 116074.
